Vibronic coupling in C$_2$H: implications for astrochemistry

### 2022-03-21 Aug2021-CH2-.tex
add ref

\subsection{Coupling Calculations}
In order to guide spectral assignment of C$_2$H and astronomical searches for other C$_n$H radicals, transition intensities are also needed. However, due to the strong vibronic coupling interactions, this can not be obtained via standard quantum chemistry methods. To account for these interactions, the photoelectron spectrum of the C$_2$H$^-$ anion was simulated using a quasidiabatic Hamiltonian of the type advocated by Koeppel, Domcke and Cederbaum (KDC, reference here).   In this approach, the Hamiltonian is represented in a basis of quasidiabatic (slowly varying) electronic  states for which the kinetic energy operator can be assumed diagonal.   For C$_2$H, the KDC Hamiltonian comprises three states - the two components of the $^2\Pi$ state and the ground $^2\Sigma$ state - and is then projected onto a vibrational basis, usually chosen as a direct product of harmonic oscillators.  Diagonalization of the corresponding matrix yields the molecular states (which are given in terms of a Born-Huang expansion), and the squared projections of the corresponding eigenvectors onto the ground state of the anion yield the relative intensities. The latter is automatically true if the photodetachment cross sections for the $^2\Sigma$ and $^2\Pi$ states are assumed equal, but different cross sections of the two states can be incorporated by scaling the intensities of states according to their (vibronic) symmetry.

Details of the construction and parametrization of KDC Hamiltonians can be found elsewhere in the literature (references), and the procedure followed will be discussed here only briefly.   The present calculations use the so-called quadratic vibronic coupling (QVC) model.  For the system at hand, the QVC model Hamiltonian  assumes the following form

\begin{align}
H &= T_n1+V\\
V &= \bordermatrix{ & X & A(a') & A(a'') \cr
	X &\substack{\Delta_0^X+F_1^Xq_1+F_3^Xq_3+\\ 
		\frac{1}{2} \sum_{ij}F^X_{ij}q_iq_j}& \lambda q_{2a} & \lambda q_{2b} \cr
	A(a') & \lambda q_{2a} & \substack{\Delta^A_0+F^A_1q_1+F^A_3q_3\\+\frac{1}{2}\sum_{ij}F^A_{ij}q_iq_j+\\\frac{1}{2}\eta(q_{2a}^2-q_{2b}^2)} & \eta q_{2a}q_{2b} \cr
	A(a'') & \lambda q_{2b} & \eta q_{2a}q_{2b} & \substack{\Delta_0^A+F_1^Aq_1+F_3^Aq_3\\+\frac{1}{2}\sum_{ij}F_{ij}^Aq_iq_j\\-\frac{1}{2}\eta(q_{2a}^2-q_{2b}^2) } } \qquad
\label{eq:calc1}
\end{align}

where the summations run over the dimensionless normal modes ($q_i$) that serve as the coordinate system for the problem.   For convenience, the latter are chosen to be those of the anion, which considerably facilitates calculation of the spectral intensities.  In Eq.~(\ref{eq:calc1}), the diagonal terms of V (excluding those that carry the Renner-Teller coupling constant $\eta$) represent the quasidiabatic potential energy surfaces of the $\Sigma$, and the two components of the $\Pi$ state, chosen here as those in which the unpaired electron lies within or is perpendicular to an arbitrarily chosen plane, designated as A(a$'$) and A(a$''$), respectively. $\Delta_0^X$ is the separation of the anion and $\Sigma$ states at the origin of the coordinate system (the vertical electron detachment energy), and $\Delta_0^{A}$ is the gap between anion and $\Pi$ states at the same geometry.

For the modes of $\sigma$ symmetry ($q_1$ and $q_3$), the diabatic forces ($F_i$) and force constants ($F_{ij}$) coincide with those of the adiabatic potential energy surfaces.   However, for the bending mode, the diabatic (F$_{22}$) and adiabatic (f$_{22}$) force constants differ, and the parametrization is somewhat more involved.   For each component of the $\Pi$ state, either the 2a or 2b component of the bending vibration will maintain the A$'$ electronic symmetry that is needed to couple with the X state.  Designating these as 2a for the A(a$'$) state and 2b for the A(a$''$) state (as is implicit in Eq.~\ref{eq:calc1}), the diabatic force constants for the bending mode in the X and A states can be written as

@@

When the potential in Eq. 1 is diagonalized, the adiabatic states that are used for its parametrization are precisely recovered through second order in displacement. To parametrize this Hamiltonian, the equation-of-motion coupled-cluster method known as EOMIP-CCSDT has been used to optimise the geometry of the ground state anion, with an Atomic Natural Orbital (ANO) basis (ANO2). Adiabatic linear and quadratic force constants are calculated from the analytical derivatives of the \emph{ab-initio} potential energy surface with respect to the anion reduced normal coordinates, at the anion geometry Q$_0$. The interstate PJT coupling constant $\lambda$ is evaluated directly using the CFOUR package to employ the quasidiabatic ansatz described in refs. This allows for direct ab-initio evaluation of the linear diabatic interstate coupling constants that parametrize the KDC potential. The calculated parameters are presented in the Supplementary Materials, in Table S1.

>>>>

\subsection{Coupling Calculations}
In order to guide spectral assignment of C$_2$H and astronomical searches for other C$_n$H radicals, transition intensities are also needed. However, due to the strong vibronic coupling interactions, this can not be obtained via standard quantum chemistry methods. To account for these interactions, the photoelectron spectrum of the C$_2$H$^-$ anion was simulated using a quasidiabatic Hamiltonian of the type advocated by Koeppel, Domcke and Cederbaum (KDC).~\cite{kou84,dom81} In this approach, the Hamiltonian is represented in a basis of quasidiabatic (slowly varying) electronic  states for which the kinetic energy operator can be assumed diagonal.~\cite{pac93} For C$_2$H, the KDC Hamiltonian comprises three states - the two components of the $^2\Pi$ state and the ground $^2\Sigma^+$ state - and is then projected onto a vibrational basis, usually chosen as a direct product of harmonic oscillators. Diagonalization of the corresponding matrix yields the molecular states (which are given in terms of a Born-Huang expansion), and the squared projections of the corresponding eigenvectors onto the ground state of the anion yield the relative intensities. The latter is automatically true if the photodetachment cross sections for the $^2\Sigma^+$ and $^2\Pi$ states are assumed equal, but different cross sections of the two states can be incorporated by scaling the intensities of states according to their (vibronic) symmetry.~\cite{sta11}

Details of the construction and parametrization of KDC Hamiltonians can be found elsewhere in the literature,~\cite{sta21,ich06,ich08,ich09,wei17} and the procedure followed will be discussed here only briefly. The present calculations use the so-called quadratic vibronic coupling (QVC) model.~\cite{ich06} For the system at hand, the QVC model Hamiltonian  assumes the following form

\begin{align}
H &= T_n1+V,\\
V &= \bordermatrix{ & X & A(a') & A(a'') \cr
	X &\substack{\Delta_0^X+F_1^Xq_1+F_3^Xq_3+\\ 
		\frac{1}{2} \sum_{ij}F^X_{ij}q_iq_j}& \lambda q_{2a} & \lambda q_{2b} \cr
	A(a') & \lambda q_{2a} & \substack{\Delta^A_0+F^A_1q_1+F^A_3q_3\\+\frac{1}{2}\sum_{ij}F^A_{ij}q_iq_j+\\\frac{1}{2}\eta(q_{2a}^2-q_{2b}^2)} & \eta q_{2a}q_{2b} \cr
	A(a'') & \lambda q_{2b} & \eta q_{2a}q_{2b} & \substack{\Delta_0^A+F_1^Aq_1+F_3^Aq_3\\+\frac{1}{2}\sum_{ij}F_{ij}^Aq_iq_j\\-\frac{1}{2}\eta(q_{2a}^2-q_{2b}^2) } } \qquad
\label{eq:calc1}
\end{align}

where the summations run over the dimensionless normal modes ($q_i$) that serve as the coordinate system for the problem.~\cite{sta21,car00} For convenience, the latter are chosen to be those of the anion, which considerably facilitates calculation of the spectral intensities.  In Eq.~(\ref{eq:calc1}), the diagonal terms of V (excluding those that carry the Renner-Teller coupling constant $\eta$) represent the quasidiabatic potential energy surfaces of the $\Sigma$, and the two components of the $\Pi$ state, chosen here as those in which the unpaired electron lies within or is perpendicular to an arbitrarily chosen plane, designated as A(a$'$) and A(a$''$), respectively. $\Delta_0^X$ is the separation of the anion and $^2\Sigma^+$ states at the origin of the coordinate system (the vertical electron detachment energy), and $\Delta_0^{A}$ is the gap between anion and $^2\Pi$ states at the same geometry.

For the modes of $\sigma$ symmetry ($q_1$ and $q_3$), the diabatic forces ($F_i$) and force constants ($F_{ij}$) coincide with those of the adiabatic potential energy surfaces.   However, for the bending mode, the diabatic (F$_{22}$) and adiabatic (f$_{22}$) force constants differ, and the parametrization is somewhat more involved.   For each component of the $\Pi$ state, either the 2a or 2b component of the bending vibration will maintain the A$'$ electronic symmetry that is needed to couple with the X state.  Designating these as 2a for the A(a$'$) state and 2b for the A(a$''$) state (as is implicit in Eq.~\ref{eq:calc1}), the diabatic force constants for the bending mode in the $^2\Sigma^+$ and $^2\Pi$ states can be written as

@@

When the potential in Eq.~(\ref{eq:calc1}) is diagonalized, the adiabatic states that are used for its parametrization are precisely recovered through terms second order in displacement. The linear diabatic coupling constants between the electronic states of interest were calculated analytically at the EOM-CCSDT/AN02 level of theory, using the CFOUR computational package and the quasidiabatic ansatz of Ichino~\emph{et al.}~\cite{ich09} Briefly, the EOM procedure is used to operationally define quasidiabatic states (those that relax according to a well-behaved reference state wavefunction) and the coupling constants are then evaluated as the first derivative of the off-diagonal elements of the electronic Hamiltonian in the basis defined by this representation. The diabatic force and coupling constants calculated to parametrize the KDC potential in Eq.~(\ref{eq:calc1}) are presented in Table~1.

%To parametrize this Hamiltonian, the equation-of-motion coupled-cluster method known as EOMIP-CCSDT has been used to optimise the geometry of the ground state anion, with an Atomic Natural Orbital (ANO) basis (ANO2), using the CFOUR computational package.

%Adiabatic linear and quadratic force constants are calculated from the analytical derivatives of the \emph{ab-initio} potential energy surface with respect to the anion reduced normal coordinates, at the anion geometry Q$_0$. The interstate PJT coupling constant $\lambda$ is evaluated directly using the CFOUR package to employ the quasidiabatic ansatz described in refs. This allows for direct ab-initio evaluation of the linear diabatic interstate coupling constants that parametrize the KDC potential. The calculated parameters are presented in the Supplementary Materials, in Table S1.
